---
title: "Insights about Pert"
author: "Dr. Jan Seifert"
date: "30 3 2021"
output: html_document
bibliography: ["pert.bib"]
biblio-style: "apalike"
link-citations: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

source("../R/threepoint_core.R")
```



## Mean

Every source says that the mean of the Pert distribution was:

$$ {\displaystyle \operatorname {E} [X]={\frac {a+4b+c}{6}}=\mu } $$

Strangely, this is also the same equation of the [double triangular distribution](http://www.mhnederlof.nl/doubletriangular.html). However, I couldn't find any reason for that and it seems like a coincidence.

Looking closer at the mathematics it becomes clear that the Pert distribution cannot guarantee the specified mean. It is a special case of the beta distribution. The general formula for the mean of the $beta(\alpha, \beta, min, max)$ distributions is ([Wikipedia](https://en.wikipedia.org/wiki/Beta_distribution#Four_parameters)):

$$
\frac\alpha{\alpha+\beta} \cdot (c-a) = \frac{\alpha c + \beta a}{\alpha+\beta}
$$

????$$\frac{2 \alpha^{\alpha} \beta^{\beta}}{B(\alpha,\beta)(\alpha + \beta)^{\alpha + \beta + 1}}(max-min)$$????

So, it is clear that it is not at all limited to the max/min parameters. It is also defined by two parameters α and β. For a long time these have defined as

$$ \alpha = {(\mu - min) * (2 \cdot mode-min-max)\over (mode-\mu)*(max - min)} $$

and

$$ \beta = \alpha \cdot {(max - \mu) \over (\mu - min)} $$


If we solve these equations for a simple example it becomes clear that there is a discrepancy. 

```{r}
min <- 1
mode <- 3
max <- 10
shape <- 4

# Compute approximate value
mu <- (min + shape * mode + max)/(shape + 2)

Params <- Pert2BetaParams(min, mode, max, shape, "classic")

# "true" mean
tMean <- Params$alpha / (Params$alpha+Params$beta) * (max-min) + min

print(Params$method)
data.frame(Arg = c("alpha", "beta", "Pert mean", "Beta mean", "Delta"),
           Val = c(Params$alpha, Params$beta, mu, tMean, mu - tMean))
```

@Davis2008 addresses this inconsistency. He proposes a new way to estimate the α and β parameters of the beta distribution to make sure that mean and standard deviation are fixed to the defined values.

$$   \alpha = \frac{2 \cdot (b + 4mode - 5a)} {3(b-a)} \cdot 
       \left[ 1 + 4 \cdot \frac{(mode-a)\cdot(b-mode)}{(b-a)^2} \right] $$

$$   \beta = \alpha \cdot \frac{5b-4mode-a}{b+4mode-5a} $$

If we make the same test with the noew equations we get:

```{r}
# min <- 1
# mode <- 3
# max <- 10
# shape <- 4
# 
# # Compute approximate value
# mu <- (min + shape * mode + max)/(shape + 2)
# 
Params <- Pert2BetaParams(min, mode, max, shape, "davis")

# "true" mean
tMean <- Params$alpha / (Params$alpha+Params$beta) * (max-min) + min

print(Params$method)
data.frame(Arg = c("alpha", "beta", "Pert mean", "Beta mean", "Delta"),
           Val = c(Params$alpha, Params$beta, mu, tMean, mu - tMean))
```



## Standard Deviation
 
The standard deviation of the Pert distribution is known as one sixth of it's total range.

$$ \sigma = {c - a \over 6} = \frac{1}{6}(c - a)$$

However, that is an approximation. The exact formula for the variance is 

$$
\sigma^2 = \frac{(m-a)(c-m)}{7}
$$

or in terms of the beta distribution:

 $$ \text{var}(X)(max-min)^2 = 
    \frac{\alpha\beta (max-min)^2} {(\alpha+\beta)^2(\alpha+\beta+1)} $$


Given the above formula of the Pert mean and the fact that the standard deviation is the square root of the variance this is the exact formula.

$$
\sigma = 
\sqrt{\frac{(\mu-a)(c-\mu)}{7}} = \\ 
\frac{1}{6} \cdot \sqrt{ \frac{5 a^2 + 16ab - 26ac - 16b^2 + 16bc + 5 c^2}{7} }
$$

Let us simplify this by setting the min $a=0$ and the max value $c=1$ for demonstration purposes. That leaves:

$$
\sigma = 
\frac{1}{6} \cdot \sqrt{ \frac{- 16b^2 + 16b + 5}{7} } = \frac{1}{6} \cdot \sqrt{ \frac{5 - 16 (b - 1) b}{7} }
$$

```{r}
library(ggplot2)
b <- seq(0, 1, 0.05)
sigma <- sqrt( (5-16*(b-1)*b) / 7 ) / 6
data <- data.frame(b, sigma)
ggplot(data, aes(x = b, y = sigma)) + geom_line()
```

As we can see the standard deviation depends on the most likely estimate $b$. It ranges around 1/6 = `r format(1/6, digits=2, nsmall = 3)` from `r format(min(sigma), digits=2, nsmall = 3)` to `r format(max(sigma), digits=2, nsmall = 3)`. Such variations are why @Farnum1987 warned about using the approximation in extreme cases when the most-likely estimate is close to the minimum or maximum, i. e. close than 13% of the total range.

Again @Davis2008 handles the inconsistency. All in all, it is probably better to avoid the `pert` distributiuon family of the [mc2d](https://rdrr.io/cran/mc2d/) (@Pouillot2010a) package. Instead, compute α and β using the equations from @Davis2008 and use the generalised [beta distribution (betagen)](https://rdrr.io/cran/mc2d/man/betagen.html) instead.


## Median

A rough estimate of the median is 

$$ Median = \frac{(a + 6*b + c)}{8} $$

The actual definition of the median is 

$$ {\displaystyle I_{\frac {1}{2}}^{[-1]}(\alpha ,\beta )(c-a)+a} $$
for which the regularized incomplete beta function 

$$ {\displaystyle I_{x}(\alpha ,\beta )={\tfrac {1}{2}}}I_x(\alpha,\beta) = \tfrac{1}{2}$$. 

However, there is no general closed-form expression for the median of the beta distribution for arbitrary values of α and β. Yet, it can easily be done in R with `qpert(0.5, min=a, mode=b, max=c)` (using the mc2d package) or with `qbeta(0.5, alpha, beta) * (c - a) + a`.


# Parametrization of the Beta distributions

## Classic



## Vose

@Vose2008 uses:

$$
\begin{aligned}
  \alpha &= \frac{(\mu - a)(2b-a-c)}{(b-\mu)(c-a)} \\
         &= \frac{(c + 4b - 5a)}{(c - a)}\\
  \beta  &= \frac{\alpha (c - \mu)}{\mu - a} \\
  \text{with } \mu &= \frac{a + 4b + c}{6}
\end{aligned}
$$


## Davis

@Davis2008 uses 

$$
\begin{aligned}
  \alpha &= \frac{2(c+4b-5a)}{3(c-a)} \cdot 
            \left[ 1+4(\frac{(b-a)(c-b)}{(c-a)^2}) \right] \\
  \beta  &= \frac{2(5c-4b-a)}{3(c-a)} \cdot 
            \left[ 1+4(\frac{(b-a)(c-b)}{(c-a)^2}) \right]
\end{aligned}
$$


# References